---
title: "Population Genetics — Conceptual Framework"
subtitle: "Lean Lecture Version"
author: 
  - name: "Peter Sørensen"
    affiliation:
      - "Center for Quantitative Genetics and Genomics"
      - "Aarhus University"
format:
  revealjs:
    theme: default
    transition: fade
    slide-number: true
    progress: true
    center: false
    auto-scale: true
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

## Population Genetics: The Big Framework

From DNA variation  
to  
evolutionary forces.

---

## What Is Population Genetics Really About?

Everything reduces to:

- Allele frequencies  
- Genotype frequencies  
- Changes across generations  

> **Evolution = change in allele frequencies over time**

---

## What Do We Actually Measure?

From genotype data we compute:

- Allele frequencies ($p$, $q$)  
- Genotype frequencies  
- Heterozygosity  
- Linkage disequilibrium  
- Genetic diversity ($\pi$, GD)

All observable patterns are summaries of frequencies.

---

## The Gene Pool

Diploid population:

- $N$ individuals  
- $2N$ alleles  

Allele frequency:

$$
p = \frac{\#A}{2N}
$$

> The gene pool is the fundamental unit of analysis.

---

## The Null Model: Hardy–Weinberg

If mating is random:

$$
f(AA)=p^2,\quad f(Aa)=2pq,\quad f(aa)=q^2
$$

$$
p^2 + 2pq + q^2 = 1
$$

This is probability — not biology.

---

## Why Hardy–Weinberg Works

Random union of gametes:

- Probability of $A$ = $p$  
- Probability of $a$ = $q$  

Multiplication rule:

$$
P(AA)=p \times p = p^2
$$

> HW assumes independence between gametes.

---

## What Hardy–Weinberg Really Means

If HW holds:

- No evolutionary forces  
- Allele frequencies constant  
- Genotype proportions predictable  

HW = **no evolution baseline**

---

## When HW Fails

Deviation implies:

- Non-random mating  
- Selection  
- Population structure  
- Drift  
- Migration  

Different forces → different patterns.

---

## Homozygote Excess: Why?

Possible explanations:

- Inbreeding  
- Wahlund effect  
- Positive assortative mating  

> Same statistical pattern, different biological causes.

---

## Inbreeding and $F$

Under inbreeding:

$$
f(Aa)=2pq(1-F)
$$

$$
f(AA)=p^2 + Fpq
$$

Inbreeding:

- Changes genotype frequencies  
- Does **not** change allele frequencies  

This distinction is critical.

---

## Genetic Drift

Finite populations:

- Sampling error each generation  
- Allele frequencies fluctuate randomly  

Drift is stronger when $N$ is small.

---

## Drift vs Selection

| Drift | Selection |
|-------|-----------|
| Random | Systematic |
| Strong in small $N$ | Depends on fitness differences |
| Can fix harmful alleles | Favors beneficial alleles |

Students must clearly distinguish these.

---

## Mutation

Mutation:

- Introduces new alleles  
- Rare per site  
- Continuous  

Without mutation:

- All variation eventually disappears.

---

## Migration (Gene Flow)

Migration:

- Moves alleles between populations  
- Reduces divergence  
- Increases within-population diversity  

Admixture leaves genomic signatures.

---

## Recombination

Recombination:

- Does not change allele frequencies  
- Changes haplotypes  
- Breaks down linkage disequilibrium  

Reshuffles genetic variation.

---

## Linkage Disequilibrium (LD)

If independent:

$$
P_{AB} = p_A p_B
$$

If not:

$$
D = P_{AB} - p_A p_B
$$

LD = non-random allele association.

---

## Why LD Exists

LD arises from:

- New mutation  
- Selection  
- Drift  
- Population structure  
- Admixture  

LD decays by recombination:

$$
D_t = (1-r)^t D_0
$$

---

## Natural Selection

Selection acts on genotypes:

$$
p' = \frac{p^2 w_{AA} + pq w_{Aa}}{\bar w}
$$

Alleles change because genotypes differ in fitness.

---

## Dominant vs Recessive Selection

- Dominant beneficial allele spreads quickly  
- Recessive beneficial allele spreads slowly when rare  

Dominance affects **rate**, not direction.

---

## Selective Sweep

When beneficial mutation rises:

- Nearby alleles hitchhike  
- Diversity drops  
- LD increases  

Sweep = reduced variation around selected site.

---

## Balancing Selection

Opposite of sweep:

- Maintains multiple alleles  
- High diversity  
- Often immune genes  

Pattern: unusually high variation.

---

## Mutation–Drift Balance

Neutral diversity:

$$
\hat{H} \approx \frac{4N_e\mu}{1+4N_e\mu}
$$

Genetic diversity reflects:

- Effective population size  
- Mutation rate  

---

## Master Summary

Population genetics connects:

Observed patterns  
→ Deviations from null  
→ Evolutionary forces  

Everything reduces to:

- Frequencies  
- Independence assumptions  
- Identifying which force caused deviation  

---

## What You Must Be Able To Do

You should be able to:

1. Calculate allele frequencies  
2. Apply Hardy–Weinberg  
3. Interpret deviations  
4. Distinguish drift vs selection  
5. Recognize sweep vs balancing selection patterns  

If you can do this — you understand the chapter.


---

# 🧮 In-Class Exercises

---

## Exercise 1 — Allele Frequencies

In a population of 200 individuals, genotype counts are:

- 60 AA  
- 100 Aa  
- 40 aa  

**Questions:**

1. What is the allele frequency of A?  
2. What is the allele frequency of a?  

Take 2 minutes. Discuss with your neighbor.

::: notes
Total individuals = 200  
Total alleles = 400  

#A alleles = 2*60 + 100 = 220  
p = 220/400 = 0.55  

q = 1 − p = 0.45  
:::

---

## Exercise 2 — Hardy–Weinberg Expectation

Suppose:

$p = 0.7$, $q = 0.3$

1. What are the expected genotype frequencies under HW?
2. If observed heterozygotes = 0.30, is there deviation?

Work it out.

::: notes
Expected:
AA = 0.49  
Aa = 0.42  
aa = 0.09  

Observed Aa = 0.30 → heterozygote deficit  

Possible causes:
- Inbreeding
- Population structure
- Selection
:::

---

## Exercise 3 — Estimating Inbreeding

Suppose:

$p = 0.6$, $q = 0.4$  
Observed heterozygotes = 0.30  

Expected under HW:

$$
2pq = 2(0.6)(0.4) = 0.48
$$

Estimate:

$$
F = 1 - \frac{H_{obs}}{H_{exp}}
$$

::: notes
F = 1 − (0.30 / 0.48)
  = 1 − 0.625
  = 0.375

Substantial inbreeding.
:::

---

## Exercise 4 — Selection Direction

Suppose:

$p = 0.4$

Fitness values:

- $w_{AA} = 1.2$
- $w_{Aa} = 1.0$
- $w_{aa} = 0.8$

Will allele A increase or decrease?

No full calculation needed — reason conceptually.

::: notes
AA has highest fitness → A favored  
A should increase  

Key point:
Selection acts on genotypes.
:::

---

# 🎯 Clicker Questions

---

## Clicker 1

Inbreeding:

A) Changes allele frequencies  
B) Changes genotype frequencies  
C) Changes both  
D) Changes neither  

::: notes
Correct: B  
Inbreeding reshuffles genotypes but does not change p.
:::

---

## Clicker 2

Which scenario produces stronger drift?

A) Large population, weak selection  
B) Small population, no selection  
C) Large population, strong selection  
D) Migration between populations  

::: notes
Correct: B  
Drift ∝ 1/N.
:::

---

## Clicker 3

A recent selective sweep produces:

A) High diversity near selected site  
B) Low diversity near selected site  
C) No change in LD  
D) Increased heterozygosity genome-wide  

::: notes
Correct: B  
Also increases LD locally.
:::

---

## Clicker 4

Recombination directly changes:

A) Allele frequencies  
B) Genotype fitness  
C) Linkage disequilibrium  
D) Mutation rate  

::: notes
Correct: C  
Recombination reshuffles alleles.
:::

---

## Clicker 5

Two subpopulations are mixed. Each is in HW individually.  
The combined population shows heterozygote deficit.

This is:

A) Selection  
B) Inbreeding  
C) Wahlund effect  
D) Drift  

::: notes
Correct: C  
Population structure causes deficit.
:::

---

## Clicker 6

A beneficial allele is recessive.

When rare, selection is:

A) Very strong  
B) Weak  
C) Neutral  
D) Impossible to determine  

::: notes
Correct: B  
Rare recessives mostly occur in heterozygotes.
:::




```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## How to use this deck

- Students watch narrated slides before class
- Class time: clickers + short calculations + discussion
- Followed by 2h exercise session

---

# Hardy–Weinberg

## HWE model

- Allele frequencies: p and q = 1 − p
- Expected genotype frequencies:
  - AA = p^2
  - Aa = 2pq
  - aa = q^2

---

## Clicker

If p = 0.7, heterozygote frequency is:

A. 0.21  
B. 0.42  
C. 0.49  
D. 0.09  

---

# LD Decay

```{r}
r <- c(0.01, 0.05, 0.2)
t <- 0:50
D0 <- 1
df <- expand.grid(t=t, r=r)
df$D <- (1-df$r)^df$t * D0
df$r <- factor(df$r)

ggplot(df, aes(t, D, color=r)) +
  geom_line(linewidth=1) +
  labs(x="Generations", y="D",
       title="LD decay: D(t) = (1-r)^t D0") +
  theme_minimal()
```

---

# Genetic Drift

```{r}
simulate_drift <- function(Ne=20, p0=0.5, gens=30, reps=20) {
  out <- lapply(seq_len(reps), function(k){
    p <- numeric(gens+1); p[1] <- p0
    for (t in 1:gens) {
      p[t+1] <- rbinom(1, 2*Ne, p[t])/(2*Ne)
    }
    data.frame(gen=0:gens, p=p, rep=k, Ne=Ne)
  })
  bind_rows(out)
}

df <- bind_rows(
  simulate_drift(Ne=20, reps=25),
  simulate_drift(Ne=200, reps=25)
)

df$Ne <- factor(df$Ne)

ggplot(df, aes(gen, p, group=rep)) +
  geom_line(alpha=0.6) +
  facet_wrap(~Ne, nrow=1) +
  labs(x="Generation", y="Allele frequency",
       title="Genetic drift: small Ne fluctuates more") +
  theme_minimal()
```

---

# Selective Sweep (Schematic)

```{r}
pos <- 1:300
center <- 150
dist <- abs(pos-center)

div <- 0.35 - 0.30*exp(-(dist^2)/(2*25^2))
ld  <- 0.10 + 0.85*exp(-(dist^2)/(2*18^2))

df1 <- data.frame(dist=dist, y=div, what="Gene diversity")
df2 <- data.frame(dist=dist, y=ld,  what="LD (r² proxy)")
df  <- bind_rows(df1, df2)

ggplot(df, aes(dist, y)) +
  geom_line(linewidth=1) +
  facet_wrap(~what, ncol=1, scales="free_y") +
  labs(x="Distance from selected site",
       title="Textbook sweep signature (schematic)") +
  theme_minimal()
```

---

# Take-home

- HWE = null model
- Inbreeding reduces heterozygosity
- LD decays with recombination
- Drift stronger in small Ne
- Selection leaves genomic signatures
